Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-15 pecomplexity.tex
finishing results and cleaning up

To further validate this finding we present an in-depth analysis of three examples (\col, \svd$_5$ and \gcc ) which nicely cover the range of complexity from low to high respectively. In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.

@@

We will start with the lowest complexity program: \col. Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{toddchaos}. If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction. This would suggest that there is some predictive structure present but not that much. However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive strucutre or equivalently the low complexity. This is in fact the case. \col is a complex signal, i.e., chaotic, but it is complex in a structured way. Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast. The existence of predictve structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.

@@

The results we just presented support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE. This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk. These findings also bring up an interesting point that this work seems to suggest. The way that structure, information and complexity is processed by a generating process plays a crucial role in the success of a given prediction scheme.

@@

In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
compare the performance of the LMA and ARIMA prediction methods (respectively)
to the value of the weighted permutation entropy for all runs of each program
under consideration. The LMA MASE values are largely similar to those of the
best predictions, primarily because LMA often performed superior to ARIMA (and
the na\"ive method). On the other hand, the ARIMA MASE values are largely
uncorrelated with WPE values. The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work. More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$). How do we determine if the prediction
strategy is not adequate for the prediction task, or if the signal is simply too
complex to predict. If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy. However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.

%By way of example, consider \col (the blue +s in
%Figure~\ref{fig:predictions_vs_wpe}). If we use the out-of-the-box ARIMA from \cite{autoArima} we get MASE ranging from $%\sim0.5$ to
%$\sim1.0$.  From that information we may assume that the best %predictions that
%can be done are twice as good as random-walk, and in some %cases predictions will
%only do as well as random-walk. In the latter such cases, the %random-walk
%prediction strategy is probably the best bet as it is so %simple and produces
%very similar error. However, if we calculate the WPE we see %that \col has a WPE
%of $0.5131 \pm 0.0034$. This incredibly low entropy implies a %great deal of
%deterministic structure which can be utilize for prediction, %even though this
%did not appear to be the case with ARIMA. In fact using LMA %on this signal we
%get an average MASE value of $0.05\pm0.0018$, an error that %is on average 20
%times better than the random walk forecast, and 10 times %better than the best
%ARIMA forcast, with little variance. Alternatively, we can %look at the other end
%of the spectrum: \gcc. This signal has a WPE of %$0.9430\pm0.0013$ and each%
%prediction method we applied was significantly out performed %by a simple random
%walk (the best strategy in the case of structureless time %series). Since \gcc
%has such little structure according to WPE it is safe to %assume that he best
%strategy will be random walk and that continuing to search %for a more accurate
%prediction strategy will likely be a fruitless process.

[[Joshua: Merge the following 2 paragraphs.]]

Figure \ref{fig:lma_vs_arima} also elucidates another fact: usable predictive structure can be present in a time series without a prediction scheme being able to utilize it. In particular, information may be transferring from past to future through the present but because of the mechanism the underlying processe uses to process that information  (e.g., linear or nonlinear) particular prediction strategies may be blind to or not be able to efficiently utilize this information. For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}. If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present as is evident in Figure \ref{fig:lma_vs_arima} (b). In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos can interpret and utilize this type of structure just fine. We believe that many of the shifts in accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that linear predictors can get at this, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.

The WPE is sensitive to both linear and nonlinear structure. When you have a low
WPE and a high ARIMA it could be that the structure WPE is picking up is simply
nonlinear structure that LMA can handle but ARIMA cannot. So while the ARIMA
prediction look really bad there is plenty of structure present as suggested by
WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it
into account and does bad.

Other noteworthy features of the LMA and ARIMA results are the cluster locations
and distributions. The WPE values of each run any particular program tend to
have little variance, leading to the clusters in
Figures~\ref{fig:arima_pred_vs_wpe}, \ref{fig:lma_pred_vs_wpe} to be fairly
constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
low too, resulting is small, tight clusters. The ARIMA MASE values of the \col
traces, however, have a large variance resulting in the spread seen in
Fig~\ref{fig:arima_pred_vs_wpe}. Not only are the MASE values of that cluster
bad, in that other predictors vastly outperform it, but they are inconsistent.
Furthermore, since LMA can predict nonlinear behavior while ARIMA can not, we
see that the clusters in Fig.~\ref{fig:arima_pred_vs_wpe} are mostly further to
the right than those in Fig.~\ref{fig:lma_pred_vs_wpe}.

%The structure being picked up on by WPE may be linear or nonlinear. When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot. So while the ARIMA prediction look realy bad there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.

%\item Talk about structure analysis. Just because there is forward information transfer, does not mean that linear predictors can get at this.

%\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by side plot.

%[[Joshua: I am not sure the following is actually important]]
%Upon closer examination of Figure \ref{fig:lma_vs_arima} many interesting features of the prediction error versus entropy landscape come to the forefront:

%\item  full results.  Image: MASE vs. WPE for both LMA \&
%ARIMA.  Points to make: 
%\begin{itemize}
%\item clusters are distributed differently
%\item clusters are shaped differently---tight or not 
%\item clusters move differently between LMA and ARIMA 
%\item finally, the diagonal line is
%important.  If you're below it, you could do better.

%\end{itemize}

%This however does not contradict our hypothesis,
%it just means that ARIMA is ill-suited for the time series under %investigation.
%To see this more clearly, consider Figure~\ref{fig:lma_vs_arima}. Since %larger
%MASE values constitute worse predictions, ARIMA performed worse than LMA in %all
%traces except {\tt SVD$_1$}.

 % section results (end)

% In Table~\ref{tab:TCMpe} \& \ref{tab:IPCpe} the behavior of the PEs as
% they limit to their asymptotic values is observed. A feature of note
% is that some of the signals drop off and some do not.  For example at
% $m=4$ {\tt col\_major} is at 0.8356, a fairly high entropy, however as
% $m$ is increased to 6 the PE plummets to 0.6880: a value consistent
% with a chaotic time series. Notice this also corresponds to a very low
% nRMSE. In contrast, consider the permutation entropy of the processor
% load of {\tt 482.sphinx3}: with $m=4$ it is 0.9951 and only drops to
% 0.9849 at $m=6$. This is consistent with the signal also having the
% highest nRMSE.

@@

@@

\section{New Figures and Tables}

>>>>

To further validate this finding we present an in-depth analysis of three examples (\col, \svdfive and \gcc ) which nicely cover the range of complexity from low to high respectively. In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.

@@

We will start with the lowest complexity program: \col. Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{mytkowicz09}. If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction. This would suggest that there is some predictive structure present but not that much. However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive structure or equivalently the low complexity. This is in fact the case. \col is a complex signal, i.e., chaotic, but it is complex in a structured way. Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast. The existence of predictive structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.

@@

[[Joshua: I am not sure if this paragraph is strong enough or even worth having]]These results support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE. This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk. The analysis of these results illuminates an interesting point: The way structure, information and complexity are processed by a generating process plays a crucial role in the success of a given prediction scheme.

@@

Comparing Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} also elucidates another key finding: usable and quantifiable predictive structure can be present in a time series without a prediction scheme being able to utilize it. In particular, information may be transferred from past to future through the present but because of the mechanism the underlying process uses to process that information  (e.g., linear or nonlinear) certain prediction strategies may be blind to or not be able to efficiently utilize this information. For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}. If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present, as is evident in Figure \ref{fig:arima_pred_vs_ent}. In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos, can interpret and utilize this type of structure just fine. We believe that many of the shifts in forecast accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that an arbitrary predictor can interpret or utilize it, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.

%In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly compare the performance of the LMA and ARIMA prediction methods (respectively) to the value of the weighted permutation entropy for all runs of each program under consideration. The LMA MASE values are largely similar to those of the best predictions, primarily because LMA often performed superior to ARIMA (and the na\"ive method). On the other hand, the ARIMA MASE values are largely uncorrelated with WPE values. As was hinted at while discussing the previous results, the fact that ARIMA is uncorrelated with WPE brings about an interesting perspective on information transfer and WPE. The WPE is sensitive to both linear and nonlinear structure. When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot. So while ARIMA is consistently out performed by random walk,  there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.

%is in large part one of the major findings of this work. More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$). How do we determine if the prediction strategy is not adequate for the prediction task, or if the signal is simply too complex to predict. If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy. However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.

Other noteworthy features of the LMA and ARIMA results are the cluster locations
and distributions. The WPE values of each run of any particular program tend to
have little variance, leading to the clusters in
Figures~\ref{fig:arima_pred_vs_ent} and \ref{fig:lma_pred_vs_ent} to be fairly
constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
low as well, resulting is small, tight clusters. The ARIMA MASE values of the \col
traces, however, have a large variance resulting in the spread seen in
Figure~\ref{fig:arima_pred_vs_ent}. Not only are the MASE values of that cluster
bad, in that other predictors vastly outperform it, but they are inconsistent.
Furthermore, since LMA can predict nonlinear behavior while ARIMA cannot, we
see that the clusters in Figure ~\ref{fig:arima_pred_vs_ent} are mostly further to
the right than those in Figure~\ref{fig:lma_pred_vs_ent}.

@@

[[Joshua: These need to be rewritten to conclude and address this paper...]]

@@
